---
title: "Two-Part Statistical Analysis of NIFM Occurrence and Mark Informativeness"
author: "Maria Cuellar"
date: "2026-03-17"
format:
  html:
    toc: true
    toc-location: body
    toc-depth: 3
    number-sections: true
    code-fold: true
execute:
  echo: true
  warning: false
  message: false
---

## Purpose

This report starts the substantive statistical analysis for the Stoney NIJ project using the most recent grouped data currently present in this project directory. The main modeling idea is a two-part or hurdle-style analysis.

That framing is natural here because the scientifically relevant outcome has two stages:

1. many tested rounds yield no observed NIFM at all
2. among rounds with a NIFM, the mark can then vary in informativeness

A single model for raw minutiae counts would mix those two processes together. For this project, it is more interpretable to separate:

1. NIFM occurrence: does a round yield any NIFM?
2. mark informativeness: conditional on a NIFM being present, how strong is that mark?

This report uses grouped counts, so the results should be treated as category-level analyses. They are useful for identifying broad patterns, but they do not yet account for clustering within guns, donors, or loads. With round-level data, the same two-part logic could be implemented more formally as an individual-level hurdle model.

## Data

The current analysis uses `data/nifm-2-12-25.csv`. Each row is a handgun type / caliber category. The `Rounds Tested` column gives the total number of rounds in that category. The remaining count columns give the number of observed NIFMs with each minutiae count.

Now load the grouped data and create the derived variables used throughout the rest of the analysis.

```{r load-and-prepare-data}
options(stringsAsFactors = FALSE)

data_candidates <- c(
  "data/nifm-2-12-25.csv",
  "../data/nifm-2-12-25.csv"
)

data_path <- data_candidates[file.exists(data_candidates)][1]
if (is.na(data_path)) {
  stop("Could not find nifm-2-12-25.csv in data/ or ../data/")
}

dat <- read.csv(data_path, check.names = FALSE)
minutiae_values <- as.numeric(names(dat)[5:ncol(dat)])
count_matrix <- as.matrix(dat[5:ncol(dat)])

dat$group <- paste0(dat$Type, dat$Caliber)
dat$total_nifm <- rowSums(count_matrix)
dat$no_nifm <- dat[["Rounds Tested"]] - dat$total_nifm
dat$occurrence_rate <- dat$total_nifm / dat[["Rounds Tested"]]
dat$mean_minutiae <- rowSums(sweep(count_matrix, 2, minutiae_values, `*`)) / dat$total_nifm
dat$nifm_ge_6 <- rowSums(count_matrix[, minutiae_values >= 6, drop = FALSE])
dat$nifm_ge_8 <- rowSums(count_matrix[, minutiae_values >= 8, drop = FALSE])
dat$nifm_ge_10 <- rowSums(count_matrix[, minutiae_values >= 10, drop = FALSE])
dat$p_ge_6 <- dat$nifm_ge_6 / dat$total_nifm
dat$p_ge_8 <- dat$nifm_ge_8 / dat$total_nifm
dat$p_ge_10 <- dat$nifm_ge_10 / dat$total_nifm

summary_tab <- dat[, c(
  "group", "Type", "Caliber", "Rounds Tested", "Donors", "total_nifm",
  "occurrence_rate", "mean_minutiae", "p_ge_6", "p_ge_8", "p_ge_10"
)]
summary_tab$occurrence_rate <- round(summary_tab$occurrence_rate, 3)
summary_tab$mean_minutiae <- round(summary_tab$mean_minutiae, 2)
summary_tab$p_ge_6 <- round(summary_tab$p_ge_6, 3)
summary_tab$p_ge_8 <- round(summary_tab$p_ge_8, 3)
summary_tab$p_ge_10 <- round(summary_tab$p_ge_10, 3)

knitr::kable(summary_tab, caption = "Grouped summary statistics by handgun type and caliber.")
```

## Why A Two-Part Model Makes Sense

At the round level, the outcome of interest is effectively zero-inflated in the ordinary-language sense: many rounds have zero observed NIFMs, while positive rounds have a right-skewed positive outcome.

For this dataset, a hurdle interpretation is more natural than a classic zero-inflated Poisson interpretation:

1. the zero state is scientifically meaningful, not just a technical nuisance
2. the positive state has its own scientific meaning because the quality of the observed NIFM matters

So the main analysis below is organized in two parts:

1. Part 1 models whether any NIFM occurs
2. Part 2 models whether the observed NIFM is more informative, using `minutiae >= 8` as the primary threshold

## Descriptive Patterns

The table above already shows two notable patterns.

1. The rate of NIFM occurrence varies widely by category, from about 23% in `R22` to about 84% in `R45`.
2. The average minutiae count among observed NIFMs is more stable, ranging from about 5.16 to 5.85, but the share of higher-minutiae NIFMs appears larger in the revolver categories `R38` and `R45`.

Now plot those two patterns directly so the separation between occurrence and informativeness is easy to see.

```{r descriptive-plots}
old_par <- par(mfrow = c(1, 2), mar = c(6, 4, 3, 1))

barplot(
  height = dat$occurrence_rate,
  names.arg = dat$group,
  ylim = c(0, 1),
  col = "steelblue",
  ylab = "Observed NIFM rate",
  main = "NIFM occurrence by group"
)

barplot(
  height = dat$p_ge_8,
  names.arg = dat$group,
  ylim = c(0, max(dat$p_ge_8) * 1.15),
  col = "tan3",
  ylab = "P(minutiae >= 8 | NIFM)",
  main = "High-minutiae share by group"
)

par(old_par)
```

## Part 1: NIFM Occurrence

For the first part of the hurdle model, the natural data structure is grouped binomial: each row gives the number of NIFMs observed out of the total number of tested rounds in that category.

### Global Heterogeneity Test

Now test whether the seven groups have the same overall NIFM occurrence rate.

```{r occurrence-heterogeneity-test}
occurrence_table <- as.matrix(dat[, c("total_nifm", "no_nifm")])
rownames(occurrence_table) <- dat$group
colnames(occurrence_table) <- c("NIFM", "No NIFM")

occurrence_chisq <- chisq.test(occurrence_table)
occurrence_chisq
```

This test strongly rejects equal occurrence rates across the seven groups.

### Group-Specific Occurrence Estimates

The simplest category-level version of Part 1 is a grouped-binomial model with one parameter per handgun-type / caliber group. This is just a convenient way to estimate the observed occurrence rates on the log-odds scale.

Now estimate the group-specific occurrence levels on the log-odds scale.

```{r occurrence-group-estimates}
occurrence_fit_group <- glm(
  cbind(total_nifm, no_nifm) ~ 0 + group,
  data = dat,
  family = binomial()
)

occurrence_coef <- data.frame(
  group = dat$group,
  occurrence_rate = round(dat$occurrence_rate, 3),
  log_odds = round(coef(occurrence_fit_group), 3)
)

knitr::kable(
  occurrence_coef,
  caption = "Group-specific occurrence estimates for Part 1 of the two-part analysis."
)
```

### Exploratory Main-Effects Grouped-Binomial Model

Because there are only seven grouped rows, any regression decomposition is exploratory. As a first pass, I fit a main-effects model with handgun type and caliber:

Now fit a simpler explanatory model that separates handgun type from caliber.

```{r occurrence-main-effects-model}
occurrence_fit <- glm(
  cbind(total_nifm, no_nifm) ~ Type + factor(Caliber),
  data = dat,
  family = quasibinomial()
)

summary(occurrence_fit)

occurrence_fitted <- data.frame(
  group = dat$group,
  observed = round(dat$occurrence_rate, 3),
  fitted = round(fitted(occurrence_fit), 3)
)

knitr::kable(
  occurrence_fitted,
  caption = "Observed and fitted NIFM occurrence rates from the exploratory grouped-binomial model."
)
```

The main conclusion from this model is not that any single coefficient is precisely estimated; with only seven grouped rows, coefficient-level inference is fragile. The useful takeaway is that the data clearly show between-group heterogeneity in Part 1 of the hurdle model.

## Part 2: Mark Informativeness Among Observed NIFMs

For the second part of the hurdle model, the question is no longer whether a NIFM exists. Instead, conditional on observing a NIFM, the question is whether some groups tend to produce more informative marks.

The raw minutiae categories are sparse in the upper tail, so I use coarser bands:

- `4_5`
- `6_7`
- `8plus`

### Three-Band Distribution Test

Now compare the full low-versus-mid-versus-high minutiae distribution across groups.

```{r minutiae-band-test}
band_table <- cbind(
  `4_5` = rowSums(count_matrix[, minutiae_values %in% c(4, 5), drop = FALSE]),
  `6_7` = rowSums(count_matrix[, minutiae_values %in% c(6, 7), drop = FALSE]),
  `8plus` = rowSums(count_matrix[, minutiae_values >= 8, drop = FALSE])
)
rownames(band_table) <- dat$group

band_chisq <- chisq.test(band_table, simulate.p.value = TRUE, B = 100000)
band_chisq

knitr::kable(
  cbind(group = rownames(band_table), as.data.frame(band_table)),
  caption = "Observed minutiae-count bands among NIFMs by group."
)
```

This provides evidence that the minutiae distribution differs across groups, although the effect is more modest than the group differences in overall NIFM occurrence.

### Focused High-Minutiae Analysis

Because David's notes emphasize the practical value of more informative marks, I also examine the probability that an observed NIFM has at least 8 minutiae.

Now focus on the more practical threshold question: among observed NIFMs, how often does a mark reach at least 8 minutiae?

```{r high-minutiae-analysis}
high_table <- cbind(
  High = dat$nifm_ge_8,
  Lower = dat$total_nifm - dat$nifm_ge_8
)
rownames(high_table) <- dat$group

high_chisq <- chisq.test(high_table)
high_chisq

high_fit_group <- glm(
  cbind(nifm_ge_8, total_nifm - nifm_ge_8) ~ 0 + group,
  data = dat,
  family = binomial()
)

high_coef <- data.frame(
  group = dat$group,
  observed_probability = round(dat$p_ge_8, 3),
  log_odds = round(coef(high_fit_group), 3)
)

knitr::kable(
  high_coef,
  caption = "Group-specific high-detail probabilities for Part 2 of the two-part analysis."
)

high_fit <- glm(
  cbind(nifm_ge_8, total_nifm - nifm_ge_8) ~ Type + factor(Caliber),
  data = dat,
  family = quasibinomial()
)

summary(high_fit)

high_fitted <- data.frame(
  group = dat$group,
  total_nifm = dat$total_nifm,
  observed = round(dat$p_ge_8, 3),
  fitted = round(fitted(high_fit), 3)
)

knitr::kable(
  high_fitted,
  caption = "Observed and fitted probabilities that an observed NIFM has at least 8 minutiae."
)
```

The simple chi-square test is suggestive but not definitive at the 0.05 level, while the exploratory main-effects model points in the same direction: revolver categories appear to yield a larger share of high-minutiae NIFMs than automatic categories, especially in `R38` and `R45`.

## Two-Part Interpretation

Taken together, the two-part results suggest that category differences operate more strongly through NIFM occurrence than through minutiae quality alone.

More specifically:

1. Part 1 is the strongest signal in the data: some categories are much more likely than others to produce any NIFM.
2. Part 2 also shows group differences, but they are more modest: among observed NIFMs, some groups appear more likely to produce marks with at least 8 minutiae.
3. This means a category can be scientifically favorable for two different reasons:
   - it yields NIFMs often
   - or, when it yields them, they tend to be more informative

That distinction is exactly why the two-part framework is better suited to this problem than a single omnibus count model.

## Interpretation

At this stage, the strongest empirical result is:

1. The categories differ substantially in how often NIFMs occur at all.

A secondary result is:

2. The categories may also differ in the quality distribution of observed NIFMs, with some revolver categories producing a larger fraction of higher-minutiae marks.

By contrast:

3. The average minutiae count alone does not appear to separate the groups as strongly as the two-part analysis does.

## Limitations

These analyses are a good starting point, but there are important limits to what can be claimed from the current file alone.

1. The data are aggregated by category, so there is no adjustment for within-gun, within-donor, or within-load dependence.
2. The absent `R32` category is a structural gap and is not modeled directly here.
3. The grouped main-effects models are exploratory because there are only seven category-level rows.
4. The current data do not yet incorporate ESLR or any direct measure of associative value.
5. Because the data are aggregated, this is a grouped approximation to a hurdle model, not the full individual-round version.

## Recommended Next Analysis Steps

The next most useful extensions are:

1. Create a round-level dataset if those underlying records exist.
2. Fit the same two-part model at the round level, with random effects for donor and gun if available.
3. Model higher-detail marks directly, for example `minutiae >= 8` or `minutiae >= 10`.
4. Incorporate ESLR or another selectivity measure once those values are available.
5. Decide whether the scientific target is best framed as:
   - occurrence of any NIFM,
   - occurrence of a sufficiently informative NIFM,
   - or total evidential yield per load.

For now, this report gives a reproducible baseline analysis that is much closer to the actual project questions than the earlier power-only draft, and it does so in the two-part framework that best matches the zero-heavy structure of the outcome.
